Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-12-16 ms.tex
3.1 ongoing2

\citet{Ho:2009a} calculates the accretion rates onto a sample of
early-type galaxies both by using X-ray observations to determine the
Bondi accretion rate, and by estimating the contributions from the
mass loss of evolved stars.  Both methods lead him to conclude that
the available reservoir of gas is more than sufficient to power
low-luminosity AGN, assuming the standard radiative efficiency $\eta
\sim 0.1$ for thin disk accretion.  Indeed, there is now strong
observational evidence that the low levels of AGN activity in
quiescent galaxies results from accretion proceeding in a radiatively
inefficient mode (\citealt{Yuan&Narayan14}), either due to the
gravitationally-released energy being advected across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to only a fraction of the gas
inflowing on large scales ultimately reaching the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Radiative cooling, if important, can qualitative alter key features of the flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ according to $\dot{q}_{\rm cool} \propto T^{-2.7}$ (eq.~[\ref{eq:cooling}] below) is thermally unstable.  Thermal instability leads to higher black hole accretion rates and enhanced star formation which, when coupled by feed-back processes, can lead to limit cycle behavior, inconsistent with our assumption of a a steady-state flow (e.g. Ciotti refs).

Radiative cooling can, however, be stabilized by the other source of (stable) wind cooling, such as the last term $\propto c_{s}^{2} \propto T$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q v_{\rm w}^{2}/2$.  To assess the conditions for thermal stability we compare the nominal rate of wind heating to the rate of radiative cooling $\dot{q}_{\rm cool} =\Lambda(T) n^2$ at the stagnation radius,
\begin{align}
\left.\frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhostarrs}) and
(\ref{eq:rhors}).  Here $n = \rho/\mu m_p$ and we have approximated the cooling function $\Lambda(T)$ assuming solar metallicity gas using piece-wise power law (\citealt{Draine:2011a}, Fig. 34.1):
\begin{align}
\Lambda(T)\simeq
  \begin{cases}
    2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}&, T > 2 \times 10^7 \text{K} \\
    1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}&, T < 2 \times 10^7 \text{K}     
  \end{cases}
  \label{eq:cooling2}
\end{align}
There thus exists a critical minimum wind velocity required for thermal stability at the stagnation radius
\begin{align}
v_{\rm crit} \simeq
  \begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
In $\S\ref{sec:numerical}$ we will show that $v_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.  Our estimate in equation (\ref{eq:cooling2}) assumes the existence of a stagnation radius according to equation (\ref{eq:rs}), the validity of which requires $v_{\rm w} \gtrsim \sigma \sim v_{\rm crit}$.

@@

This section describes the results of our numerical calculations and tests the accuracy of the analytic expressions provided in the previous section.  Numerical solutions are calculated for each galaxy for three different values of the stellar wind heating parameter $v_{w,0}$ = 300, 600, 1200 km s$^{-1}$.

@@

What is a physically motivated value of $r_{\rm heat}$?
\citet{Bromberg+11} show that a jet of luminosity $L_{\rm j}$ and half
opening angle $\theta_{\rm j} = 0.1$ (similar to those of AGN jets)
propagates through a gaseous mass $M_{\rm g}$ to a radius $r$ in a
timescale
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Writing $L_{\rm j} = \epsilon_{\rm j}\dot{M}c^{2}$ and approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$ we find that the ratio of the jet crossing time to the free-fall time $t_{\rm ff} \sim r/\sigma$ is given by

@@

>>>>

\citet{Ho:2009a} calculates the accretion rates onto a sample of
early-type galaxies both by using X-ray observations to determine the
Bondi accretion rate, and by estimating the contributions from the
mass loss of evolved stars.  Both methods lead him to conclude that
the available reservoir of gas is more than sufficient to power
low-luminosity AGN, assuming the s
If cooling becomes important in the flow ($v_{w} < v_{\rm crit}$), it may lead to thermal instabilities.  There are two kinds of ``instabilities", global and local.  The global instability is simply that, if cooling becomes very rapid, the gas loses thermal pressure and starts to inflow faster, i.e. the entire CNM accretes.  In
practice this should could manifest as a large increase in the stagnation radius, due an effective {\it decrease} in the effective heating rate.  This global ``instability" is not a true instability insofar as the gas does not necessarily clump locally into low-T clouds.

However, \citet{McCourt+12} argue that even if cooling is exactly balanced by heating everywhere, there is still a local thermal instability that operates to turn hot gas into cold clouds.  However, this local instability only causes a significant local drop-out of mass from the hot medium (to low-T clouds), if the cooling timescale,
\begin{align}
\frac{\tcool}{\tff} &\equiv \frac{3\rho k T/2 \mu \mp}{\dot{q}_{\rm cool}} \approx \frac{9}{5} \left. \frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rs}
\end{align} 
is locally shorter than the approximately 10 times the free-fall
timescale $t_{\rm ff}$, where the last equality is evaluated at the stagnation radius and makes use of equations~(\ref{eq:Tanalytic}) and (\ref{eq:rhors2}).  We thus see that the condition $v_w \gtrsim v_{\rm crit}$ is also the condition for local instability, although the prefactor may be slightly different.
tandard radiative efficiency $\eta
\sim 0.1$ for thin disk accretion.  Indeed, there is now strong
observational evidence that the low levels of AGN activity in
quiescent galaxies results from accretion proceeding in a radiatively
inefficient mode (\citealt{Yuan&Narayan14}), either due to the
gravitationally-released energy being advected across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to only a fraction of the gas
inflowing on large scales ultimately reaching the SMBH
(e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Radiative cooling, if important, qualitative alters key features of the accretion flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm cool} \propto T^{-2.7}$ (see below) is thermally unstable.  If manifest, thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Wang+14}).

Radiative cooling can, however, be stabilized by the other sources of cooling, such as the stellar wind source term $\propto -q c_{s}^{2} \propto -qT$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q v_{\rm w}^{2}/2$ (neglecting conduction and radiative cooling).  Thermal stability at the stagnation radius can thus be assessed by comparing the nominal wind heating rate to the rate of radiative cooling $|\dot{q}_{\rm cool}| =|\Lambda(T) n^2|$,
\begin{align}
\left.\frac{q \vw^2/2}{|\dot{q}_{\rm cool}|}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs}) for the gaseous and stellar densities at the stagnation radius, respectively.  Here $n = \rho/\mu m_p$ and we have approximated the cooling function for $T < 2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}$ assuming solar metallicity gas (\citealt{Draine:2011a}, Fig. 34.1).  To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3\rho kT/2\mu m_p)/\dot{q}_{\rm cool}$ to the free-fall time $t_{\rm ff}$ if one uses the equality 
\begin{align}
\rho(\rs)=\frac{3 q(\rs) \tff(\rs)}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations~\eqref{eq:mdot_analytic} and~\eqref{eq:mdot_gas}.

Thermal stability ($0.5 q \vw^2/|\dot{q}_{\rm cool}| > 1$; $t_{\rm cool} > t_{\rm ff}$) at the stagnation radius thus requires a critical minimum heating rate
\begin{align}
v_{w,0} > v_{\rm crit} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling2}
\end{align}
Our numerical results in $\S\ref{sec:numerical}$ demonstrate that $v_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.  The inclusion of $\sigma$ in equation (\ref{eq:cooling2}) results because the top two lines apply only for $v_{w} > \sigma$ since their derivation assumed the existence of a stagnation radius given by equation (\ref{eq:rs_simple}).

@@

This section describes the results of our numerical calculations, which we use to assess the validity of the analytic expressions provided in the previous section.  Numerical solutions are calculated for each galaxy for three different values of the stellar wind heating parameter $v_{w,0}$ = 300, 600, 1200 km s$^{-1}$.

@@

What is a physically motivated value of $r_{\rm heat}$?
\citet{Bromberg+11} show that a jet of luminosity $L_{\rm j}$ and half
opening angle $\theta_{\rm j} = 0.1$ (similar to those of AGN jets)
propagates through a gaseous mass $M_{\rm g}$ to a radius $r$ in a
timescale
\begin{eqnarray}
t_{\rm jet} &\sim& 4000\,{\rm yr}\left(\frac{{L}_{\rm j}}{10^{40}\,\rm erg\,s^{-1}}\right)^{-1/3}\left(\frac{r}{\rm pc}\right)^{2/3}\left(\frac{M_{\rm g}}{10^{8}M_{\odot}}\right)^{1/3} 
\end{eqnarray}
Writing $L_{\rm j} = \epsilon_{\rm j}\dot{M}c^{2}$ and approximating $M_{\rm g} \sim \dot{M}t_{\rm ff}$ we find that the ratio of the jet crossing time to the free-fall time $t_{\rm ff} \sim r/\sigma$ is 
If cooling becomes important in the flow ($v_{w} < v_{\rm crit}$), it may lead to thermal instabilities.  There are two kinds of ``instabilities", global and local.  The global instability is simply that, if cooling becomes very rapid, the gas loses thermal pressure and starts to inflow faster, i.e. the entire CNM accretes.  In
practice this should could manifest as a large increase in the stagnation radius, due an effective {\it decrease} in the effective heating rate.  This global ``instability" is not a true instability insofar as the gas does not necessarily clump locally into low-T clouds.

@@

\section{Discussion}
